R-LIME: LIME法における矩形制約と最適化

### 2024-04-10 paper/paper.tex
ICPR final version

To address this limitation of LIME,
we propose a new method called R-LIME (Ruled LIME),
which provides both the contributions of each feature to the output
and the effective scope of the explanation.
R-LIME locally approximates a complex decision boundary linearly
in a rectangular region and maximizes the region
as long as the accuracy of the linear classifier exceeds a given threshold.
The rectangular region is interpretable for users because it is
expressed as a conjunction of feature predicates.
An example of the explanation by R-LIME for a sentiment prediction model
is illustrated in \cref{fig:example-rlime}.
It is clear that users can apply the insights derived from the explanation
only to the sentences containing the word ``not''.

\section{Related Work}
\begin{figure}[tbp]
  \centering
  \def\w{2.5}
  \def\ww{5.0}
  \def\h{0.7}
  \def\hh{1.4}
  \def\hhh{2.1}
  \def\hhhh{2.8}
  \begin{tikzpicture}[scale=0.9,nodestyle/.style={rectangle,auto}]
    \node[nodestyle] (post-hoc) at (0,0){post-hoc};
    \node[nodestyle] (dependent) at (-\w,-\hh){%
      \begin{tabular}{c}
        model-dependent \\
        (DTD~\cite{montavon2017explaining}, LRP~\cite{bach2015pixel})
      \end{tabular}
    };
    \node[nodestyle] (agnostic) at (\w,-\hh){\textbf{model-agnostic}};
    \node[nodestyle] (global) at (0,-\hhhh){%
      \begin{tabular}{c}
        global \\
        (PDP~\cite{friedman2001greedy}, ALE~\cite{apley2020visualizing})
      \end{tabular}
    };
    \node[nodestyle] (local) at (\ww,-\hhhh){%
      \begin{tabular}{c}
        \textbf{local} \\
        (LIME~\cite{ribeiro2016why}, Anchor~\cite{ribeiro2018anchors}, \underline{R-LIME})
      \end{tabular}
    };
    \draw(post-hoc) -- (0,-\h);
    \draw(dependent) -- (-\w,-\h) -- (\w,-\h) -- (agnostic);
    \draw(agnostic) -- (\w,-\hhh);
    \draw(global) -- (0,-\hhh)  -- (\ww,-\hhh) -- (local);
  \end{tikzpicture}
  \caption{%
    Categorization of post-hoc explanation methods.
    We focus on \emph{model-agnostic} and \emph{local} methods,
    which explain model's local behavior using only its output.
  }\label{fig:post-hoc}
\end{figure}
In this section,
we overview existing research on post-hoc explanation methods,
which explain the behavior of black-box models already trained.
As shown in \cref{fig:post-hoc},
post-hoc methods are classified into several categories.

@@

The experiments utilized the recidivism dataset \cite{schmidt1988predicting}.
This dataset contains information on 9,549 prisoners released from
North Carolina prisons between July 1, 1979, and June 30, 1980.
The dataset includes 19 items such as race, gender,
presence of alcohol dependence, number of prior offenses and recidivism.
For this experiment,
we treated the binary classification problem of predicting
the presence or absence of recidivism (Recidivism) as the target label.
We discretized continuous features and removed missing values,
resulting in 15 features.
This problem setting can be considered as a case
where a machine learning model is introduced to decide parole for prisoners.
Since such decisions can have a significant impact on a person's life,
it is crucial for users to interpret the outputs of black-box models appropriately.

The dataset, after removing missing values,
was split into training data (7,639 instances) and test data (955 instances).
A random forest model with 50 trees was trained using the training data.
Subsequently, LIME and R-LIME explanations were generated
for two instances extracted from the test data (\cref{fig:instance}).
For R-LIME, logistic regression was used as the linear approximation model,
and the distribution $\mathcal{D}$ was a multivariate normal distribution
estimated from the training data.
The beam width was set to $B=10$,
the confidence coefficient to $1-\delta=0.95$,
and the tolerance of the KL-LUCB algorithm to $\epsilon=0.05$.
Accuracy thresholds $\tau$ were set to $\tau=0.70,0.80,0.90$.

@@

The values assigned to each attribute name represent the contribution

@@

The figures also display the 5 attributes with the highest absolute contribution.

Explanations generated by LIME (\cref{fig:A-lime,fig:B-lime}) provide insights
that attributes such as having a prior offenses (Priors),
being served for a long time in prison (MonthsServed),
and committing a crime against property (Crime against Property)
primarily contribute to the positive prediction
(prediction that the prisoner will be re-arrested).
On the other hand,
attributes like older age (Age), being married (Married),
and being of white race (Race) contribute to the negative prediction
(prediction that the prisoner will not be re-arrested).
While these LIME explanations provide valuable insights into the behavior of
the black-box model,
they do not explicitly indicate the application scope of the explanations,
leaving users unable to determine to which inmates the explanations are applicable.

In contrast, R-LIME expresses the application scope of explanations
as a conjunction of predicates.
For example, the explanation for instance A
under $\tau=0.70$ (\cref{fig:A-rlime-70}) indicates that it is applicable
only to married prisoner (Married$=$Yes).
Furthermore, the explanations include the accuracy and coverage,
allowing users to evaluate its reliability.
For example, the coverage of the explanation for instance B under $\tau=0.90$
(\cref{fig:B-rlime-90}) is 0.01\%,
indicating that the decision boundaries around instance B are complex,
making it challenging to obtain a high-accuracy linear approximation.
This allows users to discern that the application scope of this explanation is
very narrow, limiting its utility.

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations (with $\tau=0.70,0.80,0.90$) using LIME and R-LIME\@.
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{experiments/exp2/box_plot}
  \caption[Comparison of Local Accuracy between R-LIME and LIME]{%
    Comparison of local accuracy between LIME and R-LIME\@.
    R-LIME achieved higher and less variable accuracy compared to LIME
    for all values of $\tau$.
  }\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME achieved higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

@@

\subsection{Behavior Regarding Imbalanced Label Distribution}
R-LIME may generate less useful explanations
when there is bias in the distribution of black-box model outputs.
When the distribution of black-box model outputs is significantly biased
for a given accuracy threshold $\tau$
(when the ratio of the minority label is less than $1-\tau$),
the approximation region generated by R-LIME covers the entire input space,
and the learned linear classifier always outputs the majority label
(\cref{fig:imbalanced}).

@@

\subsection{Changes in Reward Distribution in Best Arm Identification}\label{sec:reward}
{%
  \renewcommand{\arraystretch}{1.1}
  \begin{table}[tbp]
    \centering
    \caption[Deviation between the estimated accuracy and the true accuracy]{%
      Deviation between the estimated accuracy and the true accuracy.
      Deviation was relatively small considering confidence level $1-\delta=0.95$.
    }\label{tab:reward}
    \input{tex/tab_exp_adv}
  \end{table}
}
For R-LIME,
the problem of selecting the rule with the highest accuracy is formulated
as the optimal arm identification problem in multi-armed bandit theory,
solved using the KL-LUCB algorithm \cite{kaufmann2013information}.
However, this algorithm assumes that the reward distribution remains constant,
while in R-LIME,
the reward distribution (accuracy of the linear approximation)
changes with every update of the approximation model after sampling.
Therefore, rewards obtained at an early stage
might influence the estimated value and deviate from the true value.

We conducted an experiment to evaluate the deviation
between the estimated accuracy and the true accuracy.
We generated explanations for 3,200 data instances sampled from the dataset,
and compared the estimated accuracy with the true accuracy.
The true accuracy was calculated based on 1,000 instances sampled
within the approximation region.
The results in \cref{tab:reward} show a mean deviation of 0.012
with a standard deviation of 0.017.
By considering the confidence level $1-\delta=0.95$,
the deviation was relatively small.
While there are concerns about the theoretical validity of using the KL-LUCB algorithm,
the results suggest that the deviation is not significant in practice.

\section{Conclusion}
Existing methods for
local model-agnostic explanations of black-box classifiers have limitations
that they do not explicitly indicate the application scope of the explanations.
To address these challenges,
we proposed R-LIME, a method that locally approximates the decision boundary
of a black-box classifier linearly and provides a rectangular approximation region,
which is interpretable for users due to being expressed as a conjunction of feature predicates.
We proposed an algorithm to
maximize coverage of the approximation region
as long as the accuracy of the linear approximation model exceeds a given threshold.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation, can be evaluated by users for its reliability and generality,
and achieves higher and less variable local accuracy compared to LIME\@.
However, we discussed the instability of behavior against imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm for our problem.

>>>>

To address this limitation of LIME,
we propose a new method called R-LIME (Ruled LIME),
which provides both the contributions of each feature to the output
and the effective scope of the explanation.
R-LIME locally approximates a complex decision boundary linearly
in a rectangular region and maximizes the region
as long as the accuracy of the linear classifier exceeds a given threshold.
The region is interpretable for users because it is
expressed as a conjunction of feature predicates.
An example of the explanation by R-LIME for a sentiment prediction model
is shown in \cref{fig:example-rlime}.
It is clear that users can apply the insights derived from the explanation
only to the sentences containing the word ``not''.

\section{Related Work}
\begin{figure}[tbp]
  \centering
  \def\w{2.5}
  \def\ww{5.0}
  \def\h{0.5}
  \def\hh{1.3}
  \def\hhh{1.8}
  \def\hhhh{2.6}
  \begin{tikzpicture}[scale=0.9,nodestyle/.style={rectangle,auto}]
    \node[nodestyle] (post-hoc) at (0,0){post-hoc};
    \node[nodestyle] (dependent) at (-\w,-\hh){%
      \begin{tabular}{c}
        model-dependent \\
        (DTD~\cite{montavon2017explaining}, LRP~\cite{bach2015pixel})
      \end{tabular}
    };
    \node[nodestyle] (agnostic) at (\w,-\hh){\textbf{model-agnostic}};
    \node[nodestyle] (global) at (0,-\hhhh){%
      \begin{tabular}{c}
        global \\
        (PDP~\cite{friedman2001greedy}, ALE~\cite{apley2020visualizing})
      \end{tabular}
    };
    \node[nodestyle] (local) at (\ww,-\hhhh){%
      \begin{tabular}{c}
        \textbf{local} \\
        (LIME~\cite{ribeiro2016why}, Anchor~\cite{ribeiro2018anchors}, \underline{R-LIME})
      \end{tabular}
    };
    \draw(post-hoc) -- (0,-\h);
    \draw(dependent) -- (-\w,-\h) -- (\w,-\h) -- (agnostic);
    \draw(agnostic) -- (\w,-\hhh);
    \draw(global) -- (0,-\hhh)  -- (\ww,-\hhh) -- (local);
  \end{tikzpicture}
  \caption{%
    Categorization of post-hoc explanation methods.
    We focus on \emph{model-agnostic} and \emph{local} methods,
    which explain model's local behavior using only its output.
  }\label{fig:post-hoc}
\end{figure}
In this section,
we overview existing research on post-hoc explanation methods,
which explain the behavior of black-box models already trained.
As shown in \cref{fig:post-hoc},
post-hoc methods are classified into several categories.

@@

We used the recidivism dataset \cite{schmidt1988predicting} for our experiments.
The dataset contains information on 9549 prisoners released from
North Carolina prisons between July 1, 1979, and June 30, 1980.
The dataset includes 19 items such as race (\emph{Race}), gender (\emph{Gender}),
presence of alcohol dependence (\emph{Alcohol}),
number of prior offenses (\emph{Priors}),
and presence of recidivism (\emph{Recidivism}).
For this experiment,
we treated the binary classification problem of predicting
the presence of recidivism (\emph{Recidivism}) as the target label.
We discretized continuous features and removed missing values,
resulting in 15 features.
This problem setting can be considered as a case
where a machine learning model is introduced to decide parole for prisoners.
Since such decisions can have a significant impact on a person's life,
it is crucial for users to interpret the outputs of black-box models appropriately.

The dataset, after removing missing values,
was split into training data (7639 instances) and test data (955 instances).
A random forest model with 50 trees was trained using the training data.
Then, LIME and R-LIME explanations were generated
for two instances extracted from the test data (\cref{fig:instance}).
For R-LIME, logistic regression was used as the linear approximation model,
and the distribution $\mathcal{D}$ was a multivariate normal distribution
estimated from the training data.
The beam width was set to $B=10$,
the confidence coefficient to $1-\delta=0.95$,
and the tolerance of the KL-LUCB algorithm to $\epsilon=0.05$.
Accuracy thresholds $\tau$ were set to $\tau=0.70,0.80,0.90$.
  {%

@@

The values assigned to each feature name represent the contribution

@@

The figures display the 5 features with the highest absolute contribution.

Explanations generated by LIME (\cref{fig:A-lime,fig:B-lime}) provide insights
that attributes such as having a prior offenses (\emph{Priors}),
being served for a long time in prison (\emph{MonthsServed}),
and committing a crime against property (\emph{Crime against Property})
primarily contribute to the positive prediction
(prediction that the prisoner will be re-arrested).
On the other hand,
attributes like older age (\emph{Age}), being married (\emph{Married}),
and being of white race (\emph{Race}) contribute to the negative prediction
(prediction that the prisoner will not be re-arrested).
While these LIME explanations provide valuable insights into the behavior of
the black-box model,
they do not explicitly indicate the application scope of the explanations,
leaving users unable to determine to which prisoners the explanations are applicable.

In contrast, R-LIME expresses the application scope of explanations
as a conjunction of feature predicates.
For example, the explanation for instance A
under $\tau=0.70$ (\cref{fig:A-rlime-70}) indicates that it is applicable
only to married prisoner (\emph{Married} $=$\emph{Yes}).
Furthermore, the explanations provide their accuracy and coverage,
allowing users to evaluate reliability of the explanations.
For example, the coverage of the explanation for instance B under $\tau=0.90$
(\cref{fig:B-rlime-90}) is 0.01\%,
indicating that the decision boundaries around instance B are complex,
making it challenging to obtain a high-accuracy linear approximation.
This allows users to discern that the application scope of this explanation is
very narrow, limiting its utility.

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations using LIME and R-LIME (with $\tau=0.70,0.80,0.90$).
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{experiments/exp2/box_plot}
  \caption[Comparison of Local Accuracy between R-LIME and LIME]{%
    Comparison of local accuracy between LIME and R-LIME\@.
    R-LIME achieved higher and less variable accuracy compared to LIME
    for all values of $\tau$.
  }\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME achieved higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

@@

\subsection{Behavior Regarding Imbalanced Label Distribution}
R-LIME may generate less useful explanations
when the distribution of black-box classifier outputs is imbalanced.
When the ratio of the minority label is less than $1-\tau$,
where $\tau$ is the threshold for the accuracy of the linear classifier,
the approximation region generated by R-LIME covers the entire input space,
and the learned linear classifier always outputs the majority label
(\cref{fig:imbalanced}).

@@

\subsection{Changes in Reward Distribution in Best Arm Identification}\label{sec:reward}
{%
  \renewcommand{\arraystretch}{1.1}
  \begin{table}[tbp]
    \centering
    \caption[Deviation between the estimated accuracy and the true accuracy]{%
      Deviation between the estimated accuracy and the true accuracy.
      Deviation was relatively small considering confidence level $1-\delta=0.95$.
    }\label{tab:reward}
    \input{tex/tab_exp_adv}
  \end{table}
}
For R-LIME,
the problem of selecting the rule with the highest accuracy is formulated
as the best arm identification problem in multi-armed bandit framework,
and solved using the KL-LUCB algorithm \cite{kaufmann2013information}.
However, this algorithm assumes that the reward distribution remains constant,
while in R-LIME,
the reward distribution (accuracy of the linear approximation)
changes with every update of the approximation model after sampling.
Therefore, rewards obtained at an early stage
might influence the estimated value and
make it deviate from the true value.

We conducted an experiment to evaluate the deviation
between the estimated accuracy and the true accuracy.
We generated explanations for 3200 data instances sampled from the dataset,
and compared the estimated accuracy with the true accuracy.
The true accuracy was calculated based on 1000 instances sampled
within the approximation region.
The results in \cref{tab:reward} show a mean deviation of 0.012
with a standard deviation of 0.017.
By considering the confidence level $1-\delta=0.95$,
the deviation was relatively small.
While there are concerns about the theoretical validity of using the KL-LUCB algorithm,
these results suggest that the deviation is not significant in practice.

\section{Conclusion}
Existing methods for
local model-agnostic explanations of black-box classifiers have limitations
that they do not explicitly indicate the application scope of the explanations.
To address these challenges,
we proposed R-LIME, a method that locally approximates the decision boundary
of a black-box classifier linearly and provides a rectangular approximation region,
which is interpretable for users due to being expressed as a conjunction of feature predicates.
We proposed an algorithm to
maximize coverage of the approximation region
as long as the accuracy of the linear approximation model exceeds a given threshold.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation, can be evaluated by users for its reliability and generality,
and achieves higher and less variable local accuracy compared to LIME\@.
Finally, we discussed the instability of behavior against imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm for our problem.
